---
title: "Shag origins"
author: "Natalie Isaksson"
date: "30/08/2021"
output: html_document
bibliography: tracking_MSP.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(root.dir = 'C:/Users/isaks/OneDrive - University of the Highlands and Islands/PhD stuff/analysis/pfirth/RLab_shags', #set own
                      warning = FALSE,
                      message = FALSE,
                      fig.align = "center",
                      dev = "png",
                      cache = TRUE)

# knitr::opts_chunk$set(warning = FALSE,message = FALSE,fig.align = "center",dev = "png",cache = TRUE)
```

```{r library, echo = FALSE, include = FALSE}
#packages needed to run this script
library(lubridate)
library(dplyr)
library(ggplot2)
#library(tidyverse) #data wrangling, dplyr, lubridate and ggplot2 are all part of it, beware masking!
#library(diveMove) #tdr specific stuff
library(ggspatial) #for scalebar
library(sp) #for SpatialPointsDataFrame function
library(sf) #for anything GIS, polygons
#library(adehabitatLT) #rediscretization
#library(adehabitatHR) #utilization distributions
library(leaflet) #for quick viz
library(tmap) #for interactive plots
library(terra) #crs function
```

```{r mapping data, echo =FALSE}
#For on laptop Natalie
scot<-st_read("C:/Users/isaks/OneDrive - University of the Highlands and Islands/PhD stuff/maps/uk/Scotland_ERI", "Scotland boundary")
scot<-scot %>% st_set_crs(27700) #it is in bng

#For on laptop Inne
#scot<-st_read("scotcoast.shp")

#For both
scot_wgs84 <- st_transform(scot, "+proj=longlat +datum=WGS84") #transform to wgs84
```

This document presents the "stages" of bird-borne tracking data in prep for how these can be incorporated into marine spatial planning. While the example is for seabirds in the context of tidal stream marine renewables, the concept is applicable to any marine animal and human developments in the seascape. The example dataset is one track from a European shag tagged on Stroma, UK, in 2012.

The idea is that since tracking data can be provided to marine spatial planners at a variety of stages in its processing it is worth investigating the properties of each of these stages and thinking about how these can most appropriately be incorporated into marine spatial planning. Based on this exploration, recommendations will be made for which stage is "best" suited for marine spatial planning, however recommendations/ideas for how to incorporate any stage of tracking data will also be given as we recognize that "best" is not always possible.

#Stage 1: "raw" points

The simplest and rawest form of tracking data are the GPS fixes themselves with biologging data, where available, attached. Details on appending TDR data to GPS data (of this bird actually) are provided in the appendix of Isaksson et al. 2021.

```{r read in raw tracking data}
#I removed your path from it Natalie, hope that also works for you (should do if you have the folder "RLab_shags" as your working directory and the file is in there?)
eush616_raw <- read.csv(file = 'eush616_raw.csv') #raw tracking dataset

colnames(eush616_raw)
```

Examination of the column names reveals a bunch of variables. ID is the "name" of the bird, in this case "EUSH_616". Datetime is the time the GPS fix (position) was taken. "Activity" in this case means dive or not, and "D" = "dry" and "W" = "wet". Variable "maxdep" may be of interest in the context of tidal stream marine renewables as it gives how far below the sea surface the dive reached (in metres). All other TDR (time at depth recorder, basically the dive sensor) measurements have been removed so as not to overcomplicate this dataset.

These tracks can now be visualized.

```{r raw plot}

rawplot<-ggplot()+
  geom_sf(data=scot_wgs84,
          fill="grey70", color="white")+
  geom_point(data = eush616_raw, aes (x=Longitude, y=Latitude, colour = activity, shape = activity, fill = activity), size = 1.8, alpha =0.8)+
  scale_colour_manual(values = c("#1B9E77", "#7570B3"),
                    name = "Location",
                    breaks = c("D", "W"),
                    labels = c("Not dive", "Dive"))+
  scale_shape_manual(values = c(21, 25),
                    name = "Location",
                    breaks = c("D", "W"),
                    labels = c("Not dive", "Dive"))+
  scale_fill_manual(values = c("#1B9E77", "#7570B3"),
                    name = "Location",
                    breaks = c("D", "W"),
                    labels = c("Not dive", "Dive"))+
  coord_sf(crs = 4326, xlim=c(-3.15, -2.82), ylim=c(58.63, 58.70), expand=FALSE)+ 
  scale_y_continuous(breaks = c(58.63, 58.66, 58.69))+ 
  scale_x_continuous(breaks = c(-3.1, -3, -2.9))+
  annotation_scale(location="bl", width_hint=0.2, style="ticks")+
  annotate("text", label = "Stroma", x = -3.116, y = 58.68, size = 3, colour = "black")+
  annotate("text", label = "Muckle Skerry", x = -2.92, y = 58.695, size = 3, colour = "black")+
  #theme_bw()+
  theme(panel.background = element_rect(fill = "transparent"), 
        plot.background = element_rect(fill = "transparent", color = NA), 
        legend.background = element_rect(fill = "transparent", size =0.5, linetype = "solid", colour = "black"),
        legend.key = element_rect(fill = "transparent"),
        axis.text=element_text(size = 10), 
        axis.title =element_text(size= 14),
        panel.grid.major=element_blank(),
        panel.grid.minor=element_blank())+
labs(x="Longitude", y="Latitude")
rawplot

```

#Stage 2: Foraging trips

Seabird tracking data is usually collected during the breeding season (as this is when birds can be caught at or near the nest for putting devices on) and during this time are so-called "central place foragers" [@Orians1979). This means they are constrained to the nest (the central place) for incubation of the egg or brooding the chick or feeding the chick, and therefore perform "foraging trips". Tracking data is therefore segmented into trips, using a colony or nest buffer. This is typically species-specific, because different seabirds travel different distances. Shags don't go far so their buffer is ca. 300m from the nest. The foraging trip is a biologically meaningful unit for seabirds, there may be equivalent units for marine mammals or fish. Basically what it does is eliminate washing, socializing, resting, incubation/brooding, chick-feeding points that are at or near the nest [@Evans2016;@Christensen-Dalsgaard2017;@Gremillet2020]. 

The track below has been rediscretized to the interval the GPS was set to take fixes to "smooth" any irregularities that come about often due to the GPS device not being able to communicate with enough satellites to get a fix, in this case 120 seconds. Fixes have also been assigned to a foraging trip, based on a 300m buffer around the nest, and anything within this buffer has been removed. Details in [@Isaksson2021].

Notice how the foraging trip dataset consists of fewer points than the raw dataset (320 vs 566) and now has an added "trip_id" column, for all the foraging trips the fixes were from (in this case just one).

```{r read in foraging trip dataset}
eush616_trip <- read.csv(file = 'eush616_trip.csv') #raw tracking dataset

colnames(eush616_trip)
```

This can also be visualized.

```{r forage trip plot}
tripplot<-ggplot()+
  geom_sf(data=scot_wgs84,
          fill="grey70", color="white")+
  geom_point(data = eush616_trip, aes (x=Longitude, y=Latitude, colour = activity, shape = activity, fill = activity), size = 1.8, alpha =0.8)+
  scale_colour_manual(values = c("#1B9E77", "#7570B3"),
                    name = "Location",
                    breaks = c("D", "W"),
                    labels = c("Not dive", "Dive"))+
  scale_shape_manual(values = c(21, 25),
                    name = "Location",
                    breaks = c("D", "W"),
                    labels = c("Not dive", "Dive"))+
  scale_fill_manual(values = c("#1B9E77", "#7570B3"),
                    name = "Location",
                    breaks = c("D", "W"),
                    labels = c("Not dive", "Dive"))+
  coord_sf(crs = 4326, xlim=c(-3.15, -2.82), ylim=c(58.63, 58.70), expand=FALSE)+ 
  scale_y_continuous(breaks = c(58.63, 58.66, 58.69))+ 
  scale_x_continuous(breaks = c(-3.1, -3, -2.9))+
  annotation_scale(location="bl", width_hint=0.2, style="ticks")+
  annotate("text", label = "Stroma", x = -3.116, y = 58.68, size = 3, colour = "black")+
  annotate("text", label = "Muckle Skerry", x = -2.92, y = 58.695, size = 3, colour = "black")+
  #theme_bw()+
  theme(panel.background = element_rect(fill = "transparent"), 
        plot.background = element_rect(fill = "transparent", color = NA), 
        legend.background = element_rect(fill = "transparent", size =0.5, linetype = "solid", colour = "black"),
        legend.key = element_rect(fill = "transparent"),
        axis.text=element_text(size = 10), 
        axis.title =element_text(size= 14),
        panel.grid.major=element_blank(),
        panel.grid.minor=element_blank())+
labs(x="Longitude", y="Latitude")
tripplot
```

#Stage 3: Foraging points

In the context of marine renewables, and in particular tidal stream marine renewables, the points where birds forage and are either on the water or under water are of interest. This is because these are the points where the birds may come into contact, either directly (collision) or not (displacement), with tidal turbines. Within the foraging trip dataset there still may be points that are on land, roosting on a skerry for instance, or be of the shag in flight. In order to identify flight in order to later filter these out, some kind of behavioral classification analysis is applied, in this case a speed/tortuosity threshold. For details, see [@Isaksson2021].

The following dataset only contains points that are at sea and are NOT flight, and based on a speed/tortuosity threshold are therefore presumed foraging. There is some uncertainty in the classification because of the high tidal current speeds in the area. This is discussed in detail in [@Isaksson2021]. 

Notice that the dataset now consists of only 236 points, differing from the foraging trip and the raw dataset.

```{r read in foraging points only dataset}

eush616_foraging <- read.csv(file = 'eush616_foraging.csv') 

colnames(eush616_foraging)

```

This can be visualized again...

```{r forage trip plot}
forageplot<-ggplot()+
  geom_sf(data=scot_wgs84,
          fill="grey70", color="white")+
  geom_point(data = eush616_foraging, aes (x=Longitude, y=Latitude, colour = activity, shape = activity, fill = activity), size = 1.8, alpha =0.8)+
  scale_colour_manual(values = c("#1B9E77", "#7570B3"),
                    name = "Location",
                    breaks = c("D", "W"),
                    labels = c("Not dive", "Dive"))+
  scale_shape_manual(values = c(21, 25),
                    name = "Location",
                    breaks = c("D", "W"),
                    labels = c("Not dive", "Dive"))+
  scale_fill_manual(values = c("#1B9E77", "#7570B3"),
                    name = "Location",
                    breaks = c("D", "W"),
                    labels = c("Not dive", "Dive"))+
  coord_sf(crs = 4326, xlim=c(-3.15, -2.82), ylim=c(58.63, 58.70), expand=FALSE)+ 
  scale_y_continuous(breaks = c(58.63, 58.66, 58.69))+ 
  scale_x_continuous(breaks = c(-3.1, -3, -2.9))+
  annotation_scale(location="bl", width_hint=0.2, style="ticks")+
  annotate("text", label = "Stroma", x = -3.116, y = 58.68, size = 3, colour = "black")+
  annotate("text", label = "Muckle Skerry", x = -2.92, y = 58.695, size = 3, colour = "black")+
  #theme_bw()+
  theme(panel.background = element_rect(fill = "transparent"), 
        plot.background = element_rect(fill = "transparent", color = NA), 
        legend.background = element_rect(fill = "transparent", size =0.5, linetype = "solid", colour = "black"),
        legend.key = element_rect(fill = "transparent"),
        axis.text=element_text(size = 10), 
        axis.title =element_text(size= 14),
        panel.grid.major=element_blank(),
        panel.grid.minor=element_blank())+
labs(x="Longitude", y="Latitude")
forageplot
```
# How many dives?
```{r how many dives}
eush616_dives<-eush616_raw
eush616_dives<-eush616_dives[eush616_dives$activity=="W",]
``` 

IS LEAFLET BETTER FOR THIS KIND OF VISUALIZATION AS YOU CAN ZOOM IN???

Just so we can also zoom in, here is the plot using tmap

```{r tmap version of forage map}
#first we need to convert it to a shapefile and use the same crs as we already have
wgs84<- terra::crs(scot_wgs84)

#c(4,3) are the coordinates
eush616_foraging.sp <- SpatialPointsDataFrame(eush616_foraging[,c(4,3)],
                    eush616_foraging,    #the R object to convert
                    proj4string = wgs84)   # assign a CRS 

#then we need to make a bounding box, similar to ggplot's xlim and ylim
bbox_shag<-bbox(eush616_foraging.sp)

bbox_shag[1] <- -3.15
bbox_shag[3] <- -2.82
bbox_shag[2] <- 58.63
bbox_shag[4] <- 58.70

#plotting the map
forageplot.tm<-tm_shape(scot_wgs84,bbox=bbox_shag)+tm_polygons()+tm_shape(eush616_foraging.sp)+tm_dots(col="activity",size=0.1,title="Location",palette="viridis")
forageplot.tm
#for viewing
tmap_mode("view")
forageplot.tm
#back to plot mode
tmap_mode("plot")
forageplot.tm
```

